### 2012-04-27 report/draft.tex
Add initial report.

\title{Aggressive Scheduling for Numerical Programs}
\author{Richard Veras, Flavio Cruz, Berkin Akin}
\date{\today}
\documentclass[12pt]{article}

\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amsfonts}   % if you want the fonts
\usepackage{amssymb}    % if you want extra symbols
\usepackage{listings}
%\usepackage{savetrees}

\begin{document}

\maketitle

\section{Introduction}

Modern compilers are able to transform relatively slow code into fast code
by performing several optimizations. The majority of such optimizations
are geared towards general purpose code that is typically full of memory operations and
branches. This makes sense, since the code that most programmers write falls into this category.

However, there are certain classes of algorithms were this approach falls short in improving
the performance of the target code. One good example is numerical code. Expert programmers
are able to take a very na\"\i ve implementation and then fine tune it to take advantage of the
micro-architecture features of the target machine. Such features include things like software
pipelining (XXX include citation here) (for improving the performance of Out-Of-Order execution),
unrolling (by knowing how big are the memory caches of the processor) and pre-fetching.
Unfortunately, compilers are unable to perform such optimizations due
to some loss of generality and also because most optimizations performed by a compiler
cannot see the forest for the trees, and thus can't take into account a complete numerical algorithm.
This problem is further emphasized by the fact that expert programmers test and tune various permutations
of their code and its scheduling of instructions, a task that compilers are usually unable to do because
they would have to guess the structure of the input data and then generate and run small instances of the problem
to find an optimal solution.

Automatic Program Generation Systems like SPIRAL~\cite{Pueschel:05} is a tool that is able to produce a system
dependent and highly optimized implementation of an operation from a very high level mathematical description. 
This is achieved though three mechanism: a language that captures the algorithmic details of an operation, a powerful
rewriting mechanism that uses that knowledge to map these algorithms to the special features of the hardware (vector
instructions and parallelism), and a backend code generator that produces a tuned implementation in a high-level language
(usually C). The choice of the last item is that it provides a means for SPIRAL to be portable by allowing the compiler to do the very
system specific task of mapping the code to machine instructions.

Since SPIRAL uses C as the target language, this comes at a cost since there is no control over register allocation and
scheduling. Scheduling in particular plays a very important role in the performance of numerical code. Even on an Out-of-Order
processor long dependency chains can drastically degrade performance. Thus the goal of our project is to incorporate scheduling,
namely Software Pipelining, into SPIRAL in a way that we can produce highly-tuned schedules for target architectures.
This way, we can still use the optimizations available in modern compilers while also performing aggressive scheduling optimizations
that only experts would be able to do, the best of both worlds.

In this project we explore the problem of scheduling code at the high-level and generate high-performance code that is comparable
to an expert. XXX

\begin{comment}
+( ) APGS aim to replace expert programmers for high-performance code.
+( ) For some classes of operations they have been very successful, but for
  others they fall short in performance.

+( ) Most expert programmers code their implementations in low-level assembly
  code and perform micro-architectural specific optimizations that production
  compilers cannot do for loss of generality.

+( ) This is further aggravated by the fact that experts test and tune various
  permutations of their code and its scheduling, a task that compilers do not
  do for the same reasoning as above.

+( ) Even on out-of-order machines scheduling at the code level is important
     to minimize the effects of long dependency chains that can cause the
     system to behave like an in-order processor.

+( ) APGS tune and optimize an implementation of an operations from the
  algorithmic level down to the low level just as an expert would, but because
  these systems rely on a high-level language like C as a target instead of
  assembly.

+( ) As a result these systems must rely on the compiler for instruction
  scheduling.

+( ) One could target assembly instead of a higher-level language but that
  defeats the purpose of having a portable system that can generate
  specialized code

+( ) How can we effectively schedule code at the high-level and generate
  high-performance code that is comparable to an expert.
\end{comment}

\subsection{Our Approach}

To approach this problem, we are directly using SPIRAL \emph{intermediate code} (called ICode)
to represent several numerical algorithms. Any optimization that is done here can benefit
lower level code.

We apply several optimizations to the ICode representation of our algorithms. For example,
for loop unrolling and loop tilling (or loop blocking), we several parameters that allows
us to set how much unrolling will be done in the inner loop and the size of the tile.
This gives us great freedom to select the best parameters that will make our final code more
performant. In addition to these, we also apply software pipelining to the original code.
We create three states, namely prologue, steady state, and epilogue, to setup, run and finish
the computation, respectively.

After these transformations, we generate C code from ICode. Of course, the final code will
have all the optimizations performed at the ICode level, however we developed a technique that
forces the target C compiler to obey the scheduling we set up during the software pipelining
transformation. Usually, C compilers tend to remove such scheduling and perform their own optimizations,
which is not what we want.

Our approach gives us several advantages. The first one is that we are able to generate fully optimized
code using software pipelining from high level description. The second and final advantage, is that
we are still able to use the nice features provided by the C compiler (specially portability)
while tuning the scheduling parameters.

\begin{comment}
+( ) Work directly in the SPIRAL intermediate code. Any optimization that is done
  here can benefit any operation expressed at a higher level of abstraction.

+( ) We use Software Pipelining in addition to loop unrolling and other
  optimizations that we can fine tune.

+( ) We generate the target code in C and use a quality production compiler.

+( ) But we instrument our code in such a way that we can force a particular
  schedule on our code.

+( ) Thus we get all of the benefits of APGS and the ability to even tune
  schedules, but we can still take advantage of the compiler for dealing with
  mapping our implementation to assembly (reg alloc and such).
\end{comment}

\subsection{Related Work}
\begin{verbatim}
+( ) OSKI
+( ) FFTW
+( ) ATLAS/PhiPack
+( ) [Yotov] discusses how ATLAS only needed fine tuning and many of the course
  adjustments could be determined analytically.
+( ) Lam's software pipelining technique
\end{verbatim}

\subsection{Our Contributions}

The contributions of this paper are the following:

\begin{description}
   \item[Software Pipelining] We developed a transform that takes a numerical algorithm and adds software pipelining support
   using three stages. We are able to mix this optimization with other optimizations such as loop unrolling and loop tilling.
   Moreover, we were able to apply a technique that disallows C compilers to perform their own scheduling to the code and at
   the same time, since we generate C code, we are able to use other C compiler optimizations. 
   \item[Evaluation] We evaluated our programs using several tunable parameters for software pipelining,
   loop unrolling and loop tilling. The results
\end{description}

\begin{verbatim}
+( ) A general technique(hack) for forcing schedules in C
+( ) An implemented set of rules in SPIRAL for tunable software pipelining.
+( ) And as a result, potentially faster code that is generated by SPIRAL that is
  comparable to an expert.
\end{verbatim}

\begin{comment}
   Given a loop described at the ICode level we want to do three things:
   0. Apply a transformation to turn it into a 3 software pipelined loops
   (prologue, steady state, and epilogue). The prologue and the epilogue
   space out computation as best as possible while preserving ordering. And
   the steady state mirrors the original loop in terms of the order of
   instructions, but the elements that each instruction is operating on are
   from different loop iterations.

   1. We want unroll all 3 of the loops. We do this because we don't want
   to deal with register allocation (this would involve implementing
   iterative-modulo scheduling), we would rather let the compiler figure
   out the allocations instead. As a bonus, we drop the loop overhead.

   2. Then we want to instrument those instructions with labels from goto
   statements, so we can force the schedule. By placing a label between
   each instruction and making the control flow unpredictable at compile
   time, the compiler is forced to preserve this schedule.
\end{comment}

\section{Details Regarding Design}

% I feel this repeats the introduction
SPIRAL is an Automatic Program Generation System (APGS) that exploits the mathematical
structure of algorithms to optimize and fine tune them. Most compilers do
a poor job of compiling mathematical/numerical algorithms since they don't take advantage
of the underlying computer architecture. SPIRAL is a feedback-driven compiler because
it explores the capabilities of the target machine and tries to find the best
implementation, comparable to the one that would be hand-written by an expert programmer.
SPIRAL is thus the perfect tool to exploit software pipelining in numeral algorithms.

SPIRAL code is used to declare high level mathematical operations which compiled gets transformed
into ICode. ICode is the intermediate representation that SPIRAL uses before the code is compiled into C code.
It represents the abstract syntax tree (AST) of the parsed program.
It is also possible to write programs in ICode. An example of ICode is presented in Fig.~\ref{fig:dswap}.
It declares a function that receives two vectors (\texttt{x} and \texttt{y}) and swaps their values.
In the example, we use unrolling (the second \texttt{loop} construct) to optimize the performance.
This program has three parameters: \texttt{datatype}, to select the datatype of the vectors;
\texttt{loop\_bound}, to select the size of both vectors; and, \texttt{unroll\_bound} to select the unrolling factor.

\lstset{basicstyle=\tiny}

\begin{figure}[ht]
\begin{lstlisting}
swap := (datatype,
            loop_bound,
            unroll_bound) -> let(

   x := var("x", TPtr(datatype)),
   y := var("y", TPtr(datatype)),
   it := var("i", TInt),
   it2 := var("ii", TInt),
   actuali := var("act", TArray(TInt, loop_bound)),
   temp := var("temp", TArray(datatype, loop_bound)),

   program(
      chain(
         func(TVoid, "swap", [x, y],
            decl(Concatenation([it],
                  getVarArrayNames("temp", unroll_bound, datatype),
                  getVarArrayNames("act", unroll_bound, TInt)),
               chain(
                  loop(it, loop_bound/unroll_bound,
                     FlattenCode(loop(it2, unroll_bound,
                        chain(
                           assign(nthVar(actuali, it2), add(it2, mul(it, unroll_bound))),
                           assign(nthVar(temp, it2), nth(x, nthVar(actuali, it2))),
                           assign(nth(x, nthVar(actuali, it2)), nth(y, nthVar(actuali, it2))),
                           assign(nth(y, nthVar(actuali, it2)), nthVar(temp, it2))
                     )).unroll())
                  )
               )
            )
         )
   )
));
\end{lstlisting}
\caption{Swap Example}
\label{fig:dswap}
\end{figure}

We can ultimately compile the example in Fig.~\ref{fig:dswap} to C code. If we compile the call to \texttt{swap(TInt, 20, 2)}
we get the C code in Fig.~\ref{fig:dswap_compiled}.

\begin{figure}[ht]
\begin{lstlisting}
void swap(int  *x, int  *y) {
    int act_0, act_1, i, temp_0, temp_1;
    for(i = 0; i <= 9; i++) {
        act_0 = (i*2);
        temp_0 = x[act_0];
        x[act_0] = y[act_0];
        y[act_0] = temp_0;
        act_1 = (1 + (i*2));
        temp_1 = x[act_1];
        x[act_1] = y[act_1];
        y[act_1] = temp_1;
    }
}
\end{lstlisting}
\label{fig:dswap_compiled}
\end{figure}

\subsection{Software Pipelining}
\begin{verbatim}
+( ) Go into the background of software pipelining. How is it done?
+( ) How do we do it in SPIRAL?
     -( ) What parameters can we tune? (how much to delay an op)
     -( ) Formally what does this look like? What is our algorithm?
\end{verbatim}

\subsection{Enforcing a Schedule}
\begin{verbatim}
+( ) How do we enforce a schedule at the C level?
+( ) What does the code look like?
+( ) What is some example assembly code?

+( ) Would this hack always work? Either way it would be ideal to have a flag
     or pragma to allow the user to choose a schedule.

\end{verbatim}

\subsection{Putting it all Together}
\begin{verbatim}
+( ) Requirements of the code before we can SP (what optimizations should we
  do/evaluate automatically?)

\end{verbatim}

\section{Experimental Setup}
\begin{verbatim}
+( ) What do we want to demonstrate? Software pipelining can improve the
     performance even on Out-of-order machines.

+( ) What machines will benefit the most from this?

+( ) What operations will benefit the most?

+( ) We want to show the breakdown of the cycles for each test and where the
     stalls were occurring.
+( ) What is our testbed? What machines do we use?

\end{verbatim}

\subsection{Simple Scalar}
\begin{verbatim}
+ What is simplescalar and why are we using it?

+ We want to be able to determine the types of systems that will benefit the
  most from this kind of scheduling.

+ What parameters will we tune?

\end{verbatim}

\section{Experimental Evaluation}
\begin{verbatim}
+ What algorithms are we using?
+ What systems are we using?

+ What does the SimpleScalare results show? What systems benefit the most?
Which ones benefit the least?

+ How can we fine tune the SP parameters to a system? This is where we vary
the delays. For example the first few instruction of a loop will most likely
be loads and the rest will be computations, so can we cluster the memory ops
and issue them well before the computations.

\end{verbatim}

\section{Surprises and Lessons Learned}
\begin{verbatim}
\end{verbatim}

\section{Conclusion}
\begin{verbatim}
+
+
+
+ Future work:
  - We have incorporated delays in our system, the question is can we use the
    results from hardware counters to determine the best values for these?
\end{verbatim}

\bibliographystyle{plain}
\bibliography{draft}
\end{document}

### 2012-04-30 report/draft.tex
Add software pipelining introduction.

@@
 Our approach gives us several advantages. The first one is that we are able to generate fully optimized
-code using software pipelining from high level description. The second and final advantage, is that
+code using software pipelining from a high level description. The second and final advantage, is that
 we are still able to use the nice features provided by the C compiler (specially portability)
@@
 \subsection{Related Work}
+
 \begin{verbatim}
@@
 \end{verbatim}
+
+Lam's software pipelining technique~\cite{Lam:04}.
 
@@
 
-\begin{verbatim}
+\begin{comment}
 +( ) A general technique(hack) for forcing schedules in C
@@
   comparable to an expert.
-\end{verbatim}
+\end{comment}
 
@@
 \subsection{Software Pipelining}
+
+Software pipelining is a coding technique that optimizes loops by overlapping instructions
+from different loop iterations in order to exploit instruction level parallelism.
+Due to this overlap we say that software pipelining is a type of out-of-order execution.
+While software pipelining is somewhat similar to hardware pipelining, the order of the executions
+is not determined by the processor but by the compiler or the programmer.
+
+With software pipelining, we can reduce
+pipeline stalls in the hardware, since the pipeline can be fed with new instructions from
+different loop iterations. This greatly increases instruction parallelism because we can
+start a new loop iteration before the past ones have completed. For this technique to be
+effective, each loop is transformed into three sections:
+
+\begin{description}
+   \item[Prologue:] We start adding several iterations to the hardware pipeline.
+   This phase ends when the last instruction of the first instruction is to be executed.
+
+   \item[Steady State:] This phase starts where the prologue ends. Most execution time is spent
+   here and this finishes when the first instruction of the last iteration is executed.
+
+   \item[Epilogue:] In this phase, the hardware pipeline executes the last remaining instructions.
+\end{description}
+
+The need to duplicate the loop code into three phases is a notorious disadvantage of software
+pipelining, because it may increase the program's code size. Moreover, since software pipelining
+is usually used in conjunction with loop unrolling, this enlarges the code even more.
+
 \begin{verbatim}

### 2012-04-30 report/draft.tex
Add new reference. Add abstract.

@@
 \date{\today}
-\documentclass[12pt]{article}
+\documentclass[10pt]{article}
 
@@
 
+\begin{abstract}
+   Certain classes of algorithms are still hard to optimize by modern compilers
+   due to the differences between architecture features.
+   Numerical algorithms, for example, perform better when taking advantage of features
+   such as software pipelining and loop unrolling. Expert programmers are usually able to
+   get the last drop of performance from the hardware, however, compilers cannot do the same
+   for loss of generality. The SPIRAL system bridges this gap by generating machine specific code
+   from high level algorithmic descriptions. We used the SPIRAL system to implement software
+   pipelining for numerical algorithms with XXX (talk about scheduling). Our results show that
+   ....
+\end{abstract}
+
 \section{Introduction}
@@
 
-However, there are certain classes of algorithms were this approach falls short in improving
+However, there are certain classes of algorithms where this approach falls short in improving
 the performance of the target code. One good example is numerical code. Expert programmers
@@
 micro-architecture features of the target machine. Such features include things like software
-pipelining (XXX include citation here) (for improving the performance of Out-Of-Order execution),
+pipelining~\cite{Rau:81} (for improving the performance of Out-Of-Order execution by interleaving loop iterations),
 unrolling (by knowing how big are the memory caches of the processor) and pre-fetching.
@@
 
+\lstset{basicstyle=\small,language=C}
+
 \begin{figure}[ht]
@@
 +( ) What is our testbed? What machines do we use?
-
 \end{verbatim}
@@
 \subsection{Simple Scalar}
+
 \begin{verbatim}
 + What is simplescalar and why are we using it?
-
 + We want to be able to determine the types of systems that will benefit the
   most from this kind of scheduling.
-
 + What parameters will we tune?
-
 \end{verbatim}
@@
 \section{Experimental Evaluation}
+
 \begin{verbatim}

### 2012-05-01 report/draft.tex
Related work.

@@
 
-\begin{verbatim}
+Much research has been done to automatically generate the best implementation
+for a given architecture using tunable parameters. These parameters are
+related to standard techniques such as loop transformations, loop tilling and loop unrolling.
+In these research systems, there is a procedure that searches for the best combination of optimizations
+that will generate the most performant implementation.
+
+A very well-known system is ATLAS \cite{Whaley00automatedempirical} (Automatically Tuned Linear Algebra Software).
+It automatically generates high performant numerical libraries known as the Basic Linear Algebra Subprograms (BLAS)~\cite{Lawson:79}
+using \emph{global search} over the optimization parameters. Many programs are generated with different combinations
+and then run on the actual hardware to determine the best values that give the best performance.
+ATLAS can also make use of hand-coded BLAS routines in the case that the best solution performs worse than the hand-coded version.
+
+FFTW3~\cite{Frigo05thedesign} is also similar to ATLAS, although FFTW3 targets the implementation of the Discrete Fourier Transform.
+FFTW3 uses the concept of a \emph{planner} that receives as input the shape of the input data (array sizes, memory layouts, etc).
+The planner returns a \emph{plan} that is able to solve the problem using data that conforms to the initial shape. The planner measures
+the execution time of different plans and selects the best one. FFTW3 results show that the best plan performs very close
+to what an expert would do.
+
+While both ATLAS and FFTW3 use the same global search principle, Yotov~\cite{Yotov05issearch} showed that it is possible to get
+similar results just by using traditional \emph{analytical models}. In his work, he modified the search engine of ATLAS to use
+a model parameter estimator that receives hardware information (L1 cache size, number of registers, etc) and outputs
+one implementation of the target algorithm that should be best suited for the input parameters. Yotov wanted to show that
+traditional compilers should be able to get similar results as ATLAS without the need to execute multiple versions of the same
+algorithm to find the best solution. He showed that model driven optimization can achieve very similar results, without the need
+to do extensive and time consuming empirical search.
+
+\begin{comment}
 +( ) OSKI
@@
 +( ) Lam's software pipelining technique
-\end{verbatim}
+\end{comment}
 
-Lam's software pipelining technique~\cite{Lam:04}.
+Software pipelining was a technique invented by Patel and Davidson for scheduling hardware pipelines. Later Rau and Glaeser~\cite{Rau:81}
+were able to implement software pipelining in a compiler. In her seminal paper, Lam~\cite{Lam:88} showed that it is possible to implement
+software pipelining without hardware support and just instruction-level parallelism. This can be accomplished by using modulo variable
+expansion and unrolling of the loop code a small number of times, a technique called \emph{modulo renaming}.
+Lam also noted that inter-iteration dependences in a loop can create cycles in the dependence graph and therefore constraint the
+scheduling of all other instructions. Lam presented some techniques that deal with this problem.
 

### 2012-05-01 report/draft.tex
added a section. will return to this after I finish other commitment.

@@
    pipelining for numerical algorithms with XXX (talk about scheduling). Our results show that
-   ....
+   by unrolling and applying software pipelining at the C code level, we can
+   improve performance by up to 30\% over unrolling alone.
 \end{abstract}
@@
 pipelining, because it may increase the program's code size. Moreover, since software pipelining
-is usually used in conjunction with loop unrolling, this enlarges the code even more.
+is usually used in conjunction with loop unrolling, this enlarges the code
+even more. While this can be an issue on processors with little to no
+instruction cache, in practice on most modern systems this is not an issue.
 
@@
 +( ) How do we do it in SPIRAL?
-     -( ) What parameters can we tune? (how much to delay an op)
-     -( ) Formally what does this look like? What is our algorithm?
 \end{verbatim}
@@
 \begin{verbatim}
-+( ) How do we enforce a schedule at the C level?
++(X) How do we enforce a schedule at the C level?
 +( ) What does the code look like?
-+( ) What is some example assembly code?
-
-+( ) Would this hack always work? Either way it would be ideal to have a flag
++(X) Would this hack always work? Either way it would be ideal to have a flag
      or pragma to allow the user to choose a schedule.
@@
 \end{verbatim}
+Most modern compilers do not preserve instruction ordering imposed at the.
+input language level, and instead rely on general purpose methods such
+as \emph{list scheduling} to order instructions. In the common case this frees
+the developer from the task of determining an efficient schedule, but in our
+case this feature serves as a roadblock. While the GCC~\cite{} provides flags for
+disabling instruction scheduling it comes at the cost of overall
+performance because we only plan to schedule loop code. Moreover, our plan
+involves using the Intel Compiler~\cite{}, thus we need a general solution for
+scheduling at the C code level.
+
+In order to enforce a schedule at the C code level we need to prevent the
+compiler from imposing its own scheduling. A compiler must preserve the
+semantics of the input code, so to take advantage of this fact we place a
+label before each instruction with a \emph{goto} nested in
+a \emph{switch-case} block. We declare the value for the \emph{switch}
+statement as a global variable set such that the code will always jump to the
+first instruction. Initially, we simply embedded each instruction in
+a \emph{switch-case} block, but both the GCC and Intel compiler would create
+multiple code paths and schedule each path individually. Figure~\ref{} shows
+an example of our method in practice.
+
+For our purposes this technique proved itself effective for enforcing a
+schedule, but in some cases the GCC would add superfluous instructions. Thus
+in the future a more general approach, such as a compiler directive, would be
+advantageous.
 

### 2012-05-01 report/draft.tex
More fixes. Add original software pipelining reference.

--- 2012-05-01 report/draft.tex
@@
    Certain classes of algorithms are still hard to optimize by modern compilers
-   due to the differences between architecture features.
+   due to the differences between computer architectures.
    Numerical algorithms, for example, perform better when taking advantage of features
    such as software pipelining and loop unrolling. Expert programmers are usually able to
-   get the last drop of performance from the hardware, however, compilers cannot do the same
-   for loss of generality. The SPIRAL system bridges this gap by generating machine specific code
+   get the last drop of performance from the hardware, however, compilers cannot perform the same
+   feat for loss of generality. The SPIRAL system bridges this gap by generating machine specific code
    from high level algorithmic descriptions. We used the SPIRAL system to implement software
@@
 
-Automatic Program Generation Systems like SPIRAL~\cite{Pueschel:05} is a tool that is able to produce a system
+Automatic Program Generation Systems like SPIRAL~\cite{Pueschel:05} are tools that produce a system
 dependent and highly optimized implementation of an operation from a very high level mathematical description. 
-This is achieved though three mechanism: a language that captures the algorithmic details of an operation, a powerful
+This is achieved though three mechanisms: a language that captures the algorithmic details of an operation, a powerful
 rewriting mechanism that uses that knowledge to map these algorithms to the special features of the hardware (vector
 instructions and parallelism), and a backend code generator that produces a tuned implementation in a high-level language
-(usually C). The choice of the last item is that it provides a means for SPIRAL to be portable by allowing the compiler to do the very
-system specific task of mapping the code to machine instructions.
+(usually C). The later component provides a means for SPIRAL to be portable by allowing modern compilers to do the task
+of transforming C code to assembly.
 
-Since SPIRAL uses C as the target language, this comes at a cost since there is no control over register allocation and
+However, because SPIRAL uses C as the target language, this comes at a cost since there is no control over register allocation and
 scheduling. Scheduling in particular plays a very important role in the performance of numerical code. Even on an Out-of-Order
-processor long dependency chains can drastically degrade performance. Thus the goal of our project is to incorporate scheduling,
+processor long dependency chains can drastically degrade performance. Thus, the goal of our project is to incorporate scheduling,
 namely Software Pipelining, into SPIRAL in a way that we can produce highly-tuned schedules for target architectures.
 This way, we can still use the optimizations available in modern compilers while also performing aggressive scheduling optimizations
-that only experts would be able to do, the best of both worlds.
+that only experts would be able to do.
 
@@
 To approach this problem, we are directly using SPIRAL \emph{intermediate code} (called ICode)
-to represent several numerical algorithms. Any optimization that is done here can benefit
+to represent several numerical algorithms. Any optimization performed at this level can benefit
 lower level code.
@@
 We apply several optimizations to the ICode representation of our algorithms. For example,
-for loop unrolling and loop tilling (or loop blocking), we several parameters that allows
+for loop unrolling and loop tilling (or loop blocking), we have several parameters that allows
 us to set how much unrolling will be done in the inner loop and the size of the tile.
@@
 We create three states, namely prologue, steady state, and epilogue, to setup, run and finish
-the computation, respectively.
+the pipelined computation, respectively.
 
 After these transformations, we generate C code from ICode. Of course, the final code will
-have all the optimizations performed at the ICode level, however we developed a technique that
-forces the target C compiler to obey the scheduling we set up during the software pipelining
+have all the optimizations performed at the ICode level, however we have developed a technique that
+forces the target C compiler to obey the scheduling we have set up during the software pipelining
 transformation. Usually, C compilers tend to remove such scheduling and perform their own optimizations,
@@
 A very well-known system is ATLAS \cite{Whaley00automatedempirical} (Automatically Tuned Linear Algebra Software).
-It automatically generates high performant numerical libraries known as the Basic Linear Algebra Subprograms (BLAS)~\cite{Lawson:79}
+It automatically generates high performant numerical libraries known as the BLAS~\cite{Lawson:79} (Basic Linear Algebra Subprograms)
 using \emph{global search} over the optimization parameters. Many programs are generated with different combinations
@@
 
-Software pipelining was a technique invented by Patel and Davidson for scheduling hardware pipelines. Later Rau and Glaeser~\cite{Rau:81}
-were able to implement software pipelining in a compiler. In her seminal paper, Lam~\cite{Lam:88} showed that it is possible to implement
+Software pipelining was a technique invented by Patel and Davidson~\cite{Patel:76} for scheduling hardware pipelines.
+Later, Rau and Glaeser~\cite{Rau:81} were able to implement software pipelining in a compiler. In her seminal paper,
+Lam~\cite{Lam:88} showed that it is possible to implement
 software pipelining without hardware support and just instruction-level parallelism. This can be accomplished by using modulo variable
@@
 Lam also noted that inter-iteration dependences in a loop can create cycles in the dependence graph and therefore constraint the
-scheduling of all other instructions. Lam presented some techniques that deal with this problem.
+scheduling of all other instructions. Lam presented some techniques to deal with this problem.
 
@@
    \item[Software Pipelining] We developed a transform that takes a numerical algorithm and adds software pipelining support
-   using three stages. We are able to mix this optimization with other optimizations such as loop unrolling and loop tilling.
-   Moreover, we were able to apply a technique that disallows C compilers to perform their own scheduling to the code and at
-   the same time, since we generate C code, we are able to use other C compiler optimizations. 
+   using three stages. We are able to use this optimization along with other optimizations such as loop unrolling and loop tilling.
+   Moreover, we were able to apply a technique that disallows C compilers to perform their own scheduling to the code. At the same
+   time, we still use all the other optimizations provided by the C compiler. 
    \item[Evaluation] We evaluated our programs using several tunable parameters for software pipelining,
-   loop unrolling and loop tilling. The results
+   loop unrolling and loop tilling. The results XXX
 \end{description}

### 2012-05-02 report/draft.tex
Fix some typos. Add more references.

@@
 
+\usepackage{url}
 \usepackage{verbatim}
@@
 start a new loop iteration before the past ones have completed. For this technique to be
-effective, each loop is transformed into three sections:
+effective, the loop is transformed into three sections:
 
@@
 \subsection{Enforcing a Schedule}
-\begin{verbatim}
+\begin{comment}
 +(X) How do we enforce a schedule at the C level?
@@
 
-\end{verbatim}
-Most modern compilers do not preserve instruction ordering imposed at the.
+\end{comment}
+
+Most modern compilers do not preserve instruction ordering imposed at the
 input language level, and instead rely on general purpose methods such
-as \emph{list scheduling} to order instructions. In the common case this frees
+as \emph{list scheduling} to order instructions. In the common case, this frees
 the developer from the task of determining an efficient schedule, but in our
-case this feature serves as a roadblock. While the GCC~\cite{} provides flags for
+case this feature serves as a roadblock. While the GCC~\cite{gccwebsite} provides flags for
 disabling instruction scheduling it comes at the cost of overall
 performance because we only plan to schedule loop code. Moreover, our plan
-involves using the Intel Compiler~\cite{}, thus we need a general solution for
+involves using the Intel Compiler~\cite{iccwebsite}, thus we need a general solution for
 scheduling at the C code level.
@@
 For our purposes this technique proved itself effective for enforcing a
-schedule, but in some cases the GCC would add superfluous instructions. Thus
+schedule, but in some cases the GCC compiler would add superfluous instructions. Thus
 in the future a more general approach, such as a compiler directive, would be
@@
 
+\section{Post-mortem and Final Notes}
+
+We propose distributing the credit as XX\% to Richard, XX\% to Flávio and XX\% to Berkin.
+XXX Richard did this, Flavio did that, Berkin did ...
+
 \bibliographystyle{plain}

### 2012-05-02 report/draft.tex
incorporated the graphs

@@
 %\usepackage{savetrees}
+\usepackage{graphicx}
 
@@
 
-\subsection{Simple Scalar}
-
-\begin{verbatim}
-+ What is simplescalar and why are we using it?
-+ We want to be able to determine the types of systems that will benefit the
-  most from this kind of scheduling.
-+ What parameters will we tune?
-\end{verbatim}
-
 \section{Experimental Evaluation}
@@
 \end{verbatim}
+
+%% I know this is ugly, but if you have the chance Flavio can you beautify
+%% this with side-by-side figures
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=.5]{gemv_intel_core_i7_960.png}
+\end{center}
+\caption{Relative speedup of GEMV on the Intel Core i7 960.}
+\label{fig:gemv_corei7}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=.5]{gemv_atom.png}
+\end{center}
+\caption{Relative speedup of GEMV on the Intel Atom.}
+\label{fig:gemv_atom}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=.5]{gemv_core2_extreme_x9650.png}
+\end{center}
+\caption{Relative speedup of GEMV on the Intel Core2 Extreme X9560.}
+\label{fig:gemv_core2}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=.5]{gemv_xeon_e5320.png}
+\end{center}
+\caption{Relative speedup of GEMV on the Intel Xeon E5320.}
+\label{fig:gemv_corei7}
+\end{figure}
 

### 2012-05-02 report/draft.tex
finished up sp in spiral

@@
 
-\begin{verbatim}
-+( ) Go into the background of software pipelining. How is it done?
-+( ) How do we do it in SPIRAL?
-\end{verbatim}
+The implementation of software pipelining in SPIRAL was straightforward, and
+involved three indexing functions each for the prologue, epilogue and steady
+state, which mapped the instructions in a fully unrolled loop to its correct
+position in an unrolled and software pipelined loop.
+
+Additionally, by targeting unrolled code we were able to bypass the need to do
+register allocation in SPIRAL. By insuring that the number of instructions in the
+pipelined loop is less than the number of registers on the system, the
+compiler was able to allocate registers with little to no spills.
 

### 2012-05-02 report/draft.tex
added

@@
 a \emph{switch-case} block, but both the GCC and Intel compiler would create
-multiple code paths and schedule each path individually. Figure~\ref{} shows
+multiple code paths and schedule each path individually. Fig.~\ref{fig:force_sched} shows
 an example of our method in practice.
@@
 advantageous.
+
+\begin{figure}[ht]
+\begin{lstlisting}
+int state=0;
+
+void test(double  *Y, double  *X) {
+    switch(state) {
+      case 1: {
+      goto label1;        }
+      case 2: {
+      goto label2;        }
+      /* .... snip .... */
+      case 26: {
+      goto label26;        }
+      case 27: {
+      goto label27;        }
+    }
+
+    label1:
+    a91 = (X + 1);
+    label2:
+    a95 = *(X);
+    /* .... snip .... */
+    label26:
+    a111 = *(a94);
+    label27:
+    a112 = (Y + 3);
+}
+
+\end{lstlisting}
+\caption{A template for enforcing an instruction ordering in C.}
+\label{fig:force_sched}
+\end{figure}
 

### 2012-05-02 report/draft.tex
added schedule template
wrote up "putting it all together"... less than pleased, need more sleep.

@@
 +(X) How do we enforce a schedule at the C level?
-+( ) What does the code look like?
++(X) What does the code look like?
 +(X) Would this hack always work? Either way it would be ideal to have a flag
@@
 \subsection{Putting it all Together}
-\begin{verbatim}
+\begin{comment}
 +( ) Requirements of the code before we can SP (what optimizations should we
   do/evaluate automatically?)
+\end{comment}
+% NOTE: in the future when we submit this somewhere the following section
+% should not be a steaming pile. shame on richard
 
-\end{verbatim}
+Through the ability to apply software pipelining on any loop in SPIRAL and a
+mechanism to enforce scheduling we can tailor a schedule of any
+operation written in ICode to a given system. Currently, our optimizations for
+tailoring software pipelining include strip-mining, reordering and unrolling the
+outer loops before we perform the actual scheduling. These few optimization
+provide a great degree of freedom with respect to tuning an implementation to
+a system.
+
+However, our current system is not without limitations. Because we are
+attempting to schedule at a high level, our ICode implementation must map to the
+underlying Instruction Set Architecture(ISA). Furthermore, our method
+restricts the quality of optimizations that the compiler can provide, thus the
+ICode must be aggressively optimized. Fortunately, SPIRAL was designed with
+the necessary facilities in place for this level of optimization.
 

### 2012-05-02 report/draft.tex
Added experimental setup.

@@
 \section{Experimental Setup}
-\begin{verbatim}
+\begin{comment}
 +( ) What do we want to demonstrate? Software pipelining can improve the
@@
 
-+( ) We want to show the breakdown of the cycles for each test and where the
-     stalls were occurring.
 +( ) What is our testbed? What machines do we use?
-\end{verbatim}
+\end{comment}
+
+The goal of our experiments is to demonstrate the effectiveness of software
+pipelining at the C code level for automatically generated numerical
+operations. Instruction scheduling is one of the few optimizations that expert
+has over SPIRAL. Thus, showing that software pipelining in addition to loop
+unrolling outperforms unrolling alone will help in bridging the gap between
+experts and APGS like SPIRAL.
+
+%% NOTE: need to explain this WAY better in the future.
+For our experiments we benchmarked multiple variants of Matrix-Vector(GEMV) on
+several systems with different microarchitectures. The variants included a
+baseline in which no loops were unrolled, two unrolled variants where the
+inner-most loop was fully unrolled and the other dimension was unrolled by
+either 1 or 4. Additionally, there were two variants that were software pipelined
+versions of the unrolled code. The two different unrollings resulted in
+completely different schedules in the software pipelined code. It is worth
+noting that we specifically chose our input sizes to fit entirely in the L1
+cache, so that performance would mostly be impacted by the number of
+instruction stalls.
+
+%% NOTE: in future, need actual parameters on these systems.
+The systems we tested on were selected for their varied scheduling
+facilities with respect to instruction window size where applicable. We chose three
+out-of-order architectures (Core i7 960, Core2 X9650, Xeon E5320) and a single
+in-order system (Intel Atom).
 
@@
 \begin{center}
-\includegraphics[scale=.5]{gemv_atom.png}
-\end{center}
-\caption{Relative speedup of GEMV on the Intel Atom.}
-\label{fig:gemv_atom}
-\end{figure}
-
-\begin{figure}[ht]
-\begin{center}
 \includegraphics[scale=.5]{gemv_core2_extreme_x9650.png}
@@
 \caption{Relative speedup of GEMV on the Intel Xeon E5320.}
-\label{fig:gemv_corei7}
+\label{fig:gemv_xeon}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[scale=.5]{gemv_atom.png}
+\end{center}
+\caption{Relative speedup of GEMV on the Intel Atom.}
+\label{fig:gemv_atom}
 \end{figure}

### 2012-05-02 report/draft.tex
added experimental evaluation.

@@
 %% NOTE: need to explain this WAY better in the future.
-For our experiments we benchmarked multiple variants of Matrix-Vector(GEMV) on
-several systems with different microarchitectures. The variants included a
-baseline in which no loops were unrolled, two unrolled variants where the
-inner-most loop was fully unrolled and the other dimension was unrolled by
-either 1 or 4. Additionally, there were two variants that were software pipelined
-versions of the unrolled code. The two different unrollings resulted in
-completely different schedules in the software pipelined code. It is worth
-noting that we specifically chose our input sizes to fit entirely in the L1
-cache, so that performance would mostly be impacted by the number of
-instruction stalls.
+For our experiments we benchmarked multiple automatically generated variants
+of Matrix-Vector(GEMV) on several systems with different
+microarchitectures. The variants included a baseline in which no loops were
+unrolled, two unrolled variants where the inner-most loop was fully unrolled
+and the other dimension was unrolled by either 1 or 4. Additionally, there
+were two variants that were software pipelined versions of the unrolled
+code. The two different unrollings resulted in completely different schedules
+in the software pipelined code. It is worth noting that we specifically chose
+our input sizes to fit entirely in the L1 cache, so that performance would
+mostly be impacted by the number of instruction stalls. Lastly, we were using
+4-way 32bit floating point vector instructions.
 
@@
 
-\begin{verbatim}
+\begin{comment}
 + What algorithms are we using?
@@
 
-\end{verbatim}
+\end{comment}
+
+As anticipated, the use of software pipelining provided a substantial
+performance gain over unrolling alone, and on the more modern system (Core i7,
+Fig.~\ref{fig:gemv_corei7}) yielded an impressive 1.8x speedup over unrolling
+alone. The other two out-of-order systems (Fig.~\ref{fig:gemv_core2} and
+Fig.~\ref{fig:gemv_xeon}) saw similar gains. The reason for performance
+improvement over simple unrolling is attributed to the fact that software
+pipelining breaks dependency chains and provides the processor's ready
+instruction window with a variety of independent instructions to choose from,
+thus reducing the risk of a dependency stall.
+
+%% NOTE: need to explain the delay thing better
+Interestingly, software pipelining did not provide as significant of a gain on
+the Intel Atom (Fig.~\ref{fig:gemv_atom}) as it did on the out-of-order
+systems. This could be attributed to the fact that we were using this
+processor in 32bit mode and thus we were limited to half the number of
+registers that are available in 64bit mode. Furthermore, our current software
+pipelining implementation does not support long delays, which might have been
+necessary on this system.
 

### 2012-05-02 report/draft.tex
Added conclusion and post mortem stuff. off to get a redbull.

@@
 \section{Surprises and Lessons Learned}
-\begin{verbatim}
-\end{verbatim}
+Initially, we were going to use a simulator to model processors and determine
+how to map scheduling and unrolling parameters to the hardware
+model. Unfortunately, we wildly underestimated the difficulty of simply
+installing the simulator.
 
 \section{Conclusion}
-\begin{verbatim}
+\begin{comment}
 +
@@
     results from hardware counters to determine the best values for these?
-\end{verbatim}
+\end{comment}
+One of main limitations of APGS over expert programmers is that there are some
+optimizations that could not be done effectively without sacrificing the level
+of abstraction that such a system provides. The main limitation is that of
+scheduling, and for computationally dense operations renders APGS useless over
+experts. In this paper we demonstrated the effectiveness of schedule
+instructions in an APGS while maintaining a high level of abstraction. We
+achieved this goal in two steps. Firstly, we provided a mechanism for enforcing
+a schedule at the C code level. Secondly, we implemented a software pipelining
+transformation in SPIRAL. Through these two steps demonstrated significant
+performance gains of an automatically generated and tuned computationally
+dense operation on a variety of systems. This enhancement to SPIRAL is general
+and can easily be extended to a large class of operations.
+
+\subsection{Future Work}
+First and foremost we wish to test a larger variety of automatically generated
+operations. In particular, operations such as Matrix-Matrix Multiply stand to
+gain significantly from our system.
+
+Additionally, for the sake of the in-order systems such as the Atom, or those
+with lengthy instruction, we plan on implementing some mechanism for delaying
+instructions for arbitrary lengths.
+
+Lastly, we wish to incorporate list scheduling in addition to software
+pipelining. We suspect that the added granularity will open up further
+opportunities for performance gains.
 
@@
 
-We propose distributing the credit as XX\% to Richard, XX\% to Flávio and XX\% to Berkin.
-XXX Richard did this, Flavio did that, Berkin did ...
+We propose distributing the credit as 33\% to Richard, 33\% to Flávio and 33\% to Berkin.
+Richard extended SPIRAL and devised the C scheduling mechanism, Flávio
+implemented the benchmarks, Berkin set up the experiments.
 

### 2012-05-02 report/draft.tex
A bunch of fixes.

@@
    from high level algorithmic descriptions. We used the SPIRAL system to implement software
-   pipelining for numerical algorithms with XXX (talk about scheduling). Our results show that
+   pipelining for numerical algorithms in order to bridge this gap. Our results show that
    by unrolling and applying software pipelining at the C code level, we can
@@
 In this project we explore the problem of scheduling code at the high-level and generate high-performance code that is comparable
-to an expert. XXX
+to an expert.
 
@@
    \item[Evaluation] We evaluated our programs using several tunable parameters for software pipelining,
-   loop unrolling and loop tilling. The results XXX
+   loop unrolling and loop tilling. The results show very good speedups over code that only uses loop unrolling.
 \end{description}
@@
 from different loop iterations in order to exploit instruction level parallelism.
-Due to this overlap we say that software pipelining is a type of out-of-order execution.
+Due to this overlap we say that software pipelining is a type of Out-Of-Order execution.
 While software pipelining is somewhat similar to hardware pipelining, the order of the executions
@@
 the developer from the task of determining an efficient schedule, but in our
-case this feature serves as a roadblock. While the GCC~\cite{gccwebsite} provides flags for
+case this feature serves as a roadblock. While the GCC~\cite{gccwebsite} compiler provides flags for
 disabling instruction scheduling it comes at the cost of overall
@@
 operation written in ICode to a given system. Currently, our optimizations for
-tailoring software pipelining include strip-mining, reordering and unrolling the
+tailoring software pipelining include loop tilling (strip-mining), reordering and unrolling the
 outer loops before we perform the actual scheduling. These few optimization
@@
 attempting to schedule at a high level, our ICode implementation must map to the
-underlying Instruction Set Architecture(ISA). Furthermore, our method
+underlying Instruction Set Architecture (ISA). Furthermore, our method
 restricts the quality of optimizations that the compiler can provide, thus the
@@
 pipelining at the C code level for automatically generated numerical
-operations. Instruction scheduling is one of the few optimizations that expert
-has over SPIRAL. Thus, showing that software pipelining in addition to loop
+operations. Instruction scheduling is one of the few optimizations that programmer experts
+have over SPIRAL. Thus, showing that software pipelining in addition to loop
 unrolling outperforms unrolling alone will help in bridging the gap between
@@
 For our experiments we benchmarked multiple automatically generated variants
-of Matrix-Vector(GEMV) on several systems with different
+of Matrix-Vector (GEMV) on several systems with different
 microarchitectures. The variants included a baseline in which no loops were
@@
 our input sizes to fit entirely in the L1 cache, so that performance would
-mostly be impacted by the number of instruction stalls. Lastly, we were using
+mostly be impacted by the number of instruction stalls. Lastly, we used
 4-way 32bit floating point vector instructions.
@@
 facilities with respect to instruction window size where applicable. We chose three
-out-of-order architectures (Core i7 960, Core2 X9650, Xeon E5320) and a single
+Out-Of-Order architectures (Core i7 960, Core2 X9650, Xeon E5320) and a single
 in-order system (Intel Atom).
@@
 Fig.~\ref{fig:gemv_corei7}) yielded an impressive 1.8x speedup over unrolling
-alone. The other two out-of-order systems (Fig.~\ref{fig:gemv_core2} and
+alone. The other two Out-Of-Order systems (Fig.~\ref{fig:gemv_core2} and
 Fig.~\ref{fig:gemv_xeon}) saw similar gains. The reason for performance
@@
 Interestingly, software pipelining did not provide as significant of a gain on
-the Intel Atom (Fig.~\ref{fig:gemv_atom}) as it did on the out-of-order
+the Intel Atom (Fig.~\ref{fig:gemv_atom}) as it did on the Out-Of-Order
 systems. This could be attributed to the fact that we were using this
@@
 
-Additionally, for the sake of the in-order systems such as the Atom, or those
+Additionally, for the sake of the In-Order systems such as the Atom, or those
 with lengthy instruction, we plan on implementing some mechanism for delaying
